A Simple Minibatch Acceptance Test for MCMC

### 2016-05-20 nips_2016.tex
several minor changes. still a few more to go

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^n$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $n$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, MCMC methods typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations: 
GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed, etc.

@@

There is another advantage when using our test for Bayesian
posterior inference. As the dataset size grows, the posterior distribution
sharpens and often concentrates to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become
stuck in one of them (see Section~\ref{ssec:gaussians}).  The samples are not exploring
the posterior parameter %vspace, and are unlikely to concentrate at a
strong local optimum. They therefore fail to accurately represent a
sample of the posterior because of inadequate search (i.e.,
mixing). We argue that it is much more natural to sample the
distribution at a higher temperature determined by the
minibatch size rather than the dataset size. This temperature is
determined in a natural way by the variance prerequisite for our
test. Starting at a high temperature (small minibatch size) which
mixes easily, one can reduce the temperature (increasing minibatch
size) until one reaches the desired target distribution temperature.
Although this approach also involves changing minibatch size, the
changes increase minibatch size monotonically, and only a few changes
are performed in the course of a sampling session.
Annealing can also be used for MAP and ML estimation because changing temperature preserves
posterior modes.

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. The $P_a$ from Equation~\ref{eq:traditional}
satisfies detailed balance, so if one samples long enough, one will arrive at a stationary
distribution matching the posterior, though a burn-in period and/or taking every $n$th sample may be
helpful.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the minibatch size; in the worst case, all $N$ data points may be needed in a single iteration.  A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties
based on concentration inequalities, but comes at the cost of more computational time.

@@

There is a separate line of MCMC work on simulating the physics of a probability distribution. By
viewing random variables as particles in a system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate the best case scenario for proposals: high
quality \emph{and} distant.  HMC methods require a full gradient computation of the posterior, which
means they also face the problem of computing the likelihood at each point. Stochastic versions of
HMC~\cite{sghmc_2014,stochastic_thermostats_2014} use a minibatch of data each iteration but require
controlling the noise from sampling to avoid divergence. Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} is a similar approach which does not use the
momentum terms in HMC.  Minibatch SGHMC and Langevin Dynamics are orthogonal to our work and are
other ways of attempting to solve the same problem of MCMC methods on big data. We can combine our
methods with these results, as we demonstrate in Section~\ref{ssec:nets}.

@@

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') (\prod_{i=1}^n p(x_i \mid
\theta'))^{\frac{N}{n}} q(\theta_t \mid \theta')}{p(\theta_t) (\prod_{i=1}^n p(x_i \mid
\theta_t))^{\frac{N}{n}} q(\theta' \mid \theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are the likelihood ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively, and that the latter uses an
$N/n$ scaling term. We assume $n \ll N$ (so $\Delta'$ is substantially faster to compute) and that
the $n$ data points $x_i$ represent a random minibatch of data.

@@

\begin{lemma}\label{lem:detailed_balance}
Let $\Delta$ be defined as in Equation~\ref{eq:deltas}. Any acceptance function $g$ such that
$g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed balance. That is, $f(\theta_t)p(\theta'
\mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where $p(\theta_y \mid \theta_x)$ is the
probability of jumping from $\theta_x$ to $\theta_y$ in our chain.
\end{lemma}

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At any iteration with a current parameter $\theta_t$ and a
candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u \sim
{\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, $X$ is symmetric about zero, so the acceptance criteria can also be
expressed as $\Delta + X>0$.

@@

With large datasets, $\Delta$ is intractable to compute. An intuitive fix is to instead use the
minibatch version $\Delta'$. The following lemma characterizes the distribution of $\Delta'$.

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly and sufficiently large, then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

@@

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$. If $X_{\rm norm}$
is exact, then we get the same MH test with $\Delta'$ as we do with the full data version $\Delta$.

@@

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}), $K$ is
    small, around 5-10, and $n$ is usually a few hundred for the Central Limit Theorem to apply.
    Using these $K$ values, we use the empirical standard deviation ${\rm std}(\Delta')$.

    \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of $X_{\rm corr}$
    from standard deconvolution techniques. These rely on the well-known fact that the Fourier
    transform of a convolution is the product of the Fourier transforms. In practice, we pre-compute
    distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to facilitate the
    sampling process.
\end{enumerate}

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, number of $\Delta'$ estimates $K$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distributions, $\{x_i\}_{i=1}^n$ values from a target distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or
point estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$ and create an empty list for $\Delta'$ estimates, $d = []$\;
    \For{$k = \{1, \ldots, K\}$}{
        Sample a random minibatch of size $m$ from the complete data\;
        Compute $\Delta'$ from this minibatch (see Equation~\ref{eq:deltas}), and add $\Delta'$ to $d$\;
    }
    Compute ${\rm std}(d)$, the standard deviation estimate of $\Delta'$ from list $d$\;
    \eIf{${\rm std}(d) \ge 1.2$}{
        Skip loop (set $\theta_{t+1}=\theta_t$) and apply variance preconditioning fix (if desired)\;
    }{
        Choose the closest $X_{\rm corr}$ distribution, and sample a value $X_c$ from it\;
        Using another random minibatch of $m$ data points, compute a \emph{new} $\Delta'$ (call it $\Delta_{\rm real}'$)\;
        \eIf{$X_c + \Delta_{\rm real}' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old one, $\theta_{t+1} = \theta_t$\;
        }
    }
}
\caption{A description of our MH test within the MCMC algorithm.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

\textbf{The Convolution}. Figure~\ref{fig:deconvolution} demonstrates
Equation~\ref{eq:deconvolution} and provides three (color-coded) examples of densities for $X_{\rm
norm}$ and $X_{\rm corr}$. The density of $X$, a logistic random variable with mean $\mu = 0$ and
scale $s=1$, is known, as are the $X_{\rm norm}$ densities for different standard deviations. The
deconvolution provides us with the $X_{\rm corr}$ distributions. Note, however, that as the standard
deviation of $X_{\rm norm}$ increases, the $X_{\rm corr}$ distribution becomes increasingly unstable
and ``bumpier'', because the logistic curve has fatter tails than Gaussians. For this reason, we cap
the standard deviation possibilities of $X_{\rm norm}$ at $\approx 1.2$ in our experiments. In our
experiments, we discretize the possible standard deviations of $X_{\rm norm}$, and we also limit the
densities to consider the range $[-10,10]$ (as shown in Figure~\ref{eq:deconvolution}).

@@

\textbf{Temperature}. One way to satisfy the variance precondition is to increase the temperature of
the target distribution. For a temperature $T>1$, the augmented target distribution would become
\begin{equation}\label{eq:log_temperature}
\log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
with an extra $1/T$ term augmented. As the temperature increases, the values of $\Delta'$ get
smaller, thus reducing variance. The flatter posterior results in easier mixing.

@@

\begin{lemma}\label{lem:theory2}
For one step sampling from $\theta_t$ to $\theta'$, if the jumping step $|\theta_t - \theta'| <
\epsilon$, and the gradient of the log-likelihood is bounded by a constant factor, i.e. $|\nabla
(\log \Pr(x_i\mid \theta))| < k$, the variance of $\Delta'$ is bounded by $4\epsilon^2 k^2 (m-\frac{m(m-1)}{N-1})$, where
$m$ is the minibatch size, and $N$ is the total data size.
\end{lemma}

@@

Theorem~\ref{thm:theory3} indicates that the difference between the target and sampled distributions
consists of two components: $\rho_i$, which is determined by the efficiency of $P'_i$, and the
$\alpha_i$s, which are determined by the error of our transition kernels. These involve competing
objectives: the larger the step size size, the more efficient the $P_i'$ but its error increases.
The reverse happens with a step size too small.

@@

We start with a simple Gaussian mixture model, borrowing an experiment from~\cite{langevin_2011}.
The parameter is 2-D, $\theta = (\theta_1,\theta_2)$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
(\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. Fixing $\theta = (0,1)$, we draw 10000
data points so that the target distribution is $p(\theta)\prod_{i=1}^{10000}p(x_i\mid \theta)$, with
the prior based on the $\theta$ generation process in Equation~\ref{eq:data_generation}. This turns
out to be a rather sharp posterior, so to make our test applicable, we apply a temperature $T=110$
to reduce ${\rm std}(\Delta')$. Taking logs, we get the new target as shown in the far left of
Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test with $m=200$. We also run this using standard full-batch MCMC
(``standard sampling'') with the standard MH test, and the method from~\cite{cutting_mh_2014}
(``adaptive sampling''). For the latter, we also use $m=200$ and increment minibatches by that
amount within an iteration if necessary. The tolerance for making a decision is $\epsilon=0.1$. To
make comparisons easier, all three use the same random walk proposer with covariance $\Sigma={\rm
diag}(0.03, 0.03)$. This is a poor proposer, but it is sufficient for our purposes as the quality of
the samples will be due to the MH test. All methods are run 10000 times to collect 10000 samples.

@@

Figure~\ref{fig:gauss_mix_1} shows three scatter plots of the resulting $\theta$ samples for the
three methods, with darker regions indicating a greater density of points. All three methods obtain
the same rough form of the posterior, so our MH test can indeed obtain the posterior just as the
other methods do.

>>>>

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^N p(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^N$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $N$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and therefore have high sample efficiency. They also map very
well onto hardware such as GPUs. In contrast, MCMC methods typically
require calculations over the full dataset to produce a new sample. A
recent result from~\cite{cutting_mh_2014} attempted to bridge the gap
between minibatch optimization and MCMC by proposing an \emph{adaptive
  minibatch} method. Their method uses a subset of the $N$ data points
for the MH test each iteration. However the number of points is not
known a priori, and is adjusted dynamically during every test. This
undermines many of the practical advantages of minibatch calculations: 
GPU kernels work best for certain minibatch sizes, memory blocks
can be reused when minibatch size is fixed, etc.

@@

There is another advantage when using our test for Bayesian posterior inference. As the dataset size
grows, the posterior distribution sharpens and often concentrates to a few sharp peaks. Samples from
this distribution will come from those peaks, and will easily become stuck in one of them. The
samples are not exploring the posterior parameter space, and are unlikely to concentrate at a strong
local optimum. They therefore fail to accurately represent a sample of the posterior because of
inadequate search (i.e., mixing). We argue that it is much more natural to sample the distribution
at a higher temperature determined by the minibatch size rather than the dataset size. Starting at a
high temperature (small minibatch size) which mixes easily, one can reduce the temperature
(increasing minibatch size) until one reaches the desired target distribution temperature.  Although
this approach also involves changing minibatch size, the changes increase minibatch size
monotonically, and only a few changes are performed in the course of a sampling session.  Annealing
can also be used for MAP and ML estimation because changing temperature preserves posterior modes.

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook} for Bayesian inference with
parameter $\theta$ and conditionally independent data, the goal is to compute the distribution
$p(\theta \mid x_1, \ldots, x_N)$.  To do so, one generates a chain of (correlated) samples
$\theta_1, \ldots, \theta_T$ for large $T$ and approximates $p$ using the sample counts.  Each
iteration $t$ has a current $\theta_t$, and a \emph{proposal distribution} $q(\theta' \mid
\theta_t)$ determines a new candidate $\theta'$. With probability $P_a$, the sample is accepted, so
$\theta_{t+1} = \theta'$. Otherwise, $\theta_{t+1} = \theta_t$. This is done by drawing a uniform
random variable $u \sim {\rm Unif}[0,1]$ and accepting if $u < P_a$. Traditionally, $P_a$ is
\begin{equation}\label{eq:traditional}
P_a = \min\left\{ 1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid \theta_t)}
\right\} = \min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right\},
\end{equation}
where $f(\theta_t)=p(\theta_t \mid x_1,\ldots,x_N)$. This $P_a$ satisfies detailed balance, so if
one samples long enough, one will arrive at a stationary distribution matching the posterior, though
a burn-in period and/or only using samples at regular intervals is often done in practice.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on a probability ratio. If the test cannot make a decision
over a certain confidence threshold, then the minibatch size is increased and the test repeats. This
process continues until a decision. The downside of this algorithm is the need to keep incrementing
the minibatch size; in the worst case, all $N$ data points may be needed to decide on $\theta'$. A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties
based on concentration inequalities, but comes at the cost of more computational time.

@@

There is a separate line of MCMC work on simulating the physics of a probability distribution. By
viewing random variables as particles in a system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate the best case scenario for proposals: high
quality \emph{and} distant.  HMC methods require a full gradient computation of the posterior, which
means they also face the problem of computing the likelihood at each point. Stochastic versions of
HMC~\cite{sghmc_2014,stochastic_thermostats_2014} use a minibatch of data each iteration but require
controlling the minibatch noise to avoid divergence. Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} is a similar approach which does not use the
momentum terms in HMC.  Minibatch SGHMC and Langevin Dynamics are orthogonal to our work and are
other ways of attempting to solve the same problem of MCMC methods on big data. We can combine our
methods with these results, as we demonstrate in Section~\ref{ssec:nets}.

@@

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') (\prod_{i=1}^n p(x_i \mid
\theta'))^{\frac{N}{n}} q(\theta_t \mid \theta')}{p(\theta_t) (\prod_{i=1}^n p(x_i \mid
\theta_t))^{\frac{N}{n}} q(\theta' \mid \theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are probability ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively, and that the latter uses an
$N/n$ scaling term. We assume $n \ll N$ (so $\Delta'$ is substantially faster to compute) and that
the $n$ data points $x_i$ represent a random minibatch without replacement.

@@

\begin{lemma}\label{lem:detailed_balance}
An acceptance function $g$ such that $g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed
balance. That is, $f(\theta_t)p(\theta' \mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where
$p(\theta_y \mid \theta_x)$ is the probability of jumping from $\theta_x$ to $\theta_y$ in our
chain.
\end{lemma}

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At any iteration with a current parameter $\theta_t$ and a
candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u \sim
{\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, the density of $X$ (a standard logistic random variable) is
symmetric about zero, so the acceptance criteria can also be expressed as $\Delta + X>0$.

@@

With large datasets, $\Delta$ is intractable to compute. An intuitive fix is to replace $\Delta$
with the minibatch version, $\Delta'$, for our MH test. The following lemma characterizes the distribution of $\Delta'$.

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly without replacement and has sufficiently many elements,
then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

@@

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The noise means we cannot just replace $\Delta+X$ with
$\Delta'+X$ for our test, so we need to slightly change our acceptance criteria. Our insight is to
decompose $X$ as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ is a zero-mean Gaussian and $X_{\rm corr}$ is a zero-mean
``correction'' term.  These two add (i.e., their distributions convolve) to form $X$. The criteria
to accept is now
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists because $X_{\rm norm}$ is an estimate of $\varepsilon$. If $X_{\rm norm}$
is exact, then we get the same MH test with $\Delta'$ as we do with the full data version $\Delta$.

@@

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}) we make $K$
    small, around 5-10, for speed purposes, and $n$ is usually a few hundred for the Central Limit
    Theorem to apply.  Using these $K$ values, we compute the empirical standard deviation ${\rm
    std}(\Delta')$.

    \item Once we have ${\rm std}(\Delta')$, we can determine the distribution of $X_{\rm corr}$
    from standard deconvolution techniques. These rely on the well-known fact that the Fourier
    transform of a convolution is the product of the individual Fourier transforms. In practice, we
    pre-compute distributions of $X_{\rm corr}$ for different ${\rm std}(\Delta')$ values to
    facilitate the sampling process.
\end{enumerate}

@@

\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\begin{algorithm}[t]
\Input{Number of samples $T$, number of $\Delta'$ estimates $K$, minibatch size $m$, pre-computed $X_{\rm
corr}$ distributions, $\{x_i\}_{i=1}^N$ values from a target distribution, and initial sample $\theta_0$.}
\Output{A chain of $T$ samples $\{\theta_1, \ldots, \theta_T\}$ for either posterior inference or
MAP/ML estimation.}
\For{$t=\{1, \ldots, T\}$}{
    Propose a candidate $\theta'$ and create an empty list for $\Delta'$ estimates, $d = []$\;
    \For{$k = \{1, \ldots, K\}$}{
        Sample a random minibatch of size $m$ from the complete data\;
        Compute $\Delta'$ from this minibatch (see Equation~\ref{eq:deltas}), and add $\Delta'$ to $d$\;
    }
    Compute ${\rm std}(d)$, the standard deviation estimate of $\Delta'$ from list $d$\;
    \eIf{${\rm std}(d) \ge 1.2$}{
        Skip loop (set $\theta_{t+1}=\theta_t$) and apply variance preconditioning fix (if desired)\;
    }{
        Choose the closest $X_{\rm corr}$ distribution, and sample a value $X_c$ from it\;
        Using another random minibatch of $m$ data points, compute a \emph{new} $\Delta'$ (call it $\Delta_{\rm real}'$)\;
        \eIf{$X_c + \Delta_{\rm real}' > 0$}{
            Accept the candidate, $\theta_{t+1} = \theta'$\;
        }{
            Reject and re-use the old one, $\theta_{t+1} = \theta_t$\;
        }
    }
}
\caption{A description of our MH test within the MCMC algorithm.}
\label{alg:our_algorithm}
%\vspace{-10pt}
\end{algorithm}

\textbf{The Convolution}. Figure~\ref{fig:deconvolution} demonstrates
Equation~\ref{eq:deconvolution} and provides three (color-coded) examples of densities for $X_{\rm
norm}$ and $X_{\rm corr}$. The density of $X$, a logistic random variable with mean $\mu = 0$ and
scale $s=1$, is known, as are the $X_{\rm norm}$ densities for different standard deviations. The
deconvolution provides us with the $X_{\rm corr}$ distributions. Note, however, that as the standard
deviation of $X_{\rm norm}$ increases, the $X_{\rm corr}$ distribution becomes increasingly unstable
and ``bumpier'', because the logistic curve has fatter tails than Gaussians. In addition, we cap
the standard deviation possibilities of $X_{\rm norm}$ at $\approx 1.2$ in our experiments. In our
experiments, we discretize the possible standard deviations of $X_{\rm norm}$, and we also limit the
densities to consider the range $[-10,10]$ as shown in Figure~\ref{fig:deconvolution}.

@@

\textbf{Temperature}. One way to satisfy the variance precondition is to increase the temperature of
the target distribution. For a temperature $T>1$, the augmented target distribution would become
\begin{equation}\label{eq:log_temperature}
\log p(\theta \mid x_1,\ldots,x_N) \approx \log p(\theta) + \frac{1}{T}\frac{N}{n} \sum_{i=1}^n\log p(x_i \mid \theta).
\end{equation}
with an extra $1/T$ term augmented. As the temperature increases, the values of $\Delta'$ get
smaller in absolute value, thus reducing variance. The flatter posterior results in easier mixing.

@@

\begin{lemma}\label{lem:theory2}
For one step sampling from $\theta_t$ to $\theta'$, if the jumping step $\|\theta_t - \theta'\|_2 <
\epsilon$, and the gradient of the log likelihood is bounded by a constant factor, i.e. $\|\nabla
(\log \Pr(x_i\mid \theta))\|_2 < k$, the variance of $\Delta'$ is bounded by $4\epsilon^2 k^2
(m-\frac{m(m-1)}{N-1})$, where $m$ is the minibatch size, and $N$ is the total data size.
\end{lemma}

@@

Theorem~\ref{thm:theory3} indicates that the difference between the target and sampled distributions
consists of two components: $\rho_i$, which is determined by the efficiency of $P'_i$, and the
$\alpha_i$s, which are determined by the error of our transition kernels. These involve competing
objectives: the larger the step size, the more efficient the $P_i'$ but its error increases.  The
reverse happens with a step size too small.

@@

We start with a simple Gaussian mixture model, borrowing an experiment from~\cite{langevin_2011}.
The parameter is 2-D, $\theta = (\theta_1,\theta_2)$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
(\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. Fixing $\theta = (0,1)$, we draw 10000
data points so that the target distribution is $p(\theta)\prod_{i=1}^{10000}p(x_i\mid \theta)$, with
the prior based on the $\theta$ generation process in Equation~\ref{eq:data_generation}. This
results in rather sharp posterior modes and high ${\rm std}(\Delta')$ estimates, so we apply a
temperature $T=110$ to reduce ${\rm std}(\Delta')$. Taking logs, we get the target as shown in the
far left of Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test with $m=200$. We also run this using standard full-batch MCMC
(``standard sampling'') with the standard MH test, and the method from~\cite{cutting_mh_2014}
(``adaptive sampling''). For the latter, we also use $m=200$ and increment minibatches by that
amount within an iteration if necessary. The tolerance for making a decision is $\epsilon=0.1$. To
make comparisons easier, all three use the same random walk proposer with covariance $\Sigma={\rm
diag}(0.03, 0.03)$. This is a poor proposer, but it is sufficient for our purposes as the quality of
the samples will be primarily due to the MH test. All methods are run 10000 times to collect 10000
samples.

@@

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$ samples for the three
methods, with darker regions indicating a greater density of points. All three methods obtain the
same rough form of the posterior, so our MH test can indeed result in the same posterior as the
other two methods.
